The Power of Two Random Choices

### 2014-03-30 presentation/presentation.tex
polishing slides

@@

%---------------------------------------------------------------------
%parameters for balls into bins graphs
%---------------------------------------------------------------------
%scale factor
\newcommand\scalefac{0.55}
%diameter of balls
\newcommand\ballsize{5mm}
%number of bins
\newcommand\nrbins{6}
%space between ball and bin
\newcommand\padding{0.1*\ballsize}
%height of a bin
\newcommand\binheight{5*\balldiameter}
%width of a bin
\newcommand\binwidth{\balldiameter}
%gap from bin to bin
\newcommand\bingap{1.6*\balldiameter}
%color of balls
\tikzstyle{ballstyle} = [ball color=black!30!red]

@@

%paint a ball in bin #1, position #2
\newcommand\ball[2]{
	\shade[ballstyle] (#1*\bingap,#2*\balldiameter-\ballsize) circle (\ballsize) {};
}

@@

% nicer tikz qtrees
\tikzset{edge from parent/.style=
     {draw, edge from parent path={(\tikzparentnode) -- (\tikzchildnode)}}}
\tikzset{every tree node/.style={font=\tiny, thick}}

%---------------------------------------------------------------------
%Front matter
%---------------------------------------------------------------------
\title{The Power of Two Random Choices}
\subtitle{TODO}
\author[M. Kalany]{Martin Kalany}
\institute[TU Wien]
{
  Graduate student in Computer Science\\
  Vienna University of Technology\\
}
\date{\today}

%---------------------------------------------------------------------
%document
%---------------------------------------------------------------------
% Structure:
% 30 min
% 5 min easy stuff (so that everyone understands)
% 5 min more advanced stuff 
% 15 min difficult stuff (the proof, of the asymmetric allocation scheme using the infamous assumption 1)
% 5 min further results and wrap up
\begin{document}
\begin{frame}
  \titlepage
\end{frame}
\note{}

\begin{frame}
\frametitle{The Problem}
\begin{exampleblock}{Balls-into-bins games}
Suppose we have $n$ initially empty bins and need to distribute $m \geq n$ balls among them, \alert{as evenly as possible}. 
\end{exampleblock}

\begin{exampleblock}{Definition}
We are interested in the \alert{maximum load $\load$}, i.e, how many balls are in the fullest bin?
\end{exampleblock}

@@

\begin{frame}
\frametitle{Some definitions}
\todo{merge with next slide?}
Since we deal with a \alert{random process}, we cannot give upper bounds for the maximum load $\load$ with absolute certainty.

@@

\bigskip
%\pause

We distinguish:
\begin{itemize}
\item The \alert{lightly loaded} case: $m = n$
\item The \alert{heavily loaded} case:	 $m \in \omega(n\ln n)$
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Single choice strategy}
\framesubtitle{Upper bound for the maximum load}
\begin{theorem}[Raab and Steger, 1998]
When using the single choice strategy to distribute $m$ balls to $n$ bins, the maximum load will be, w.h.p.,
\begin{align*}
\loadS \leq 
	\begin{cases}
    \frac{\ln n}{\ln\ln n} + O(1)              & m = n \\
    \frac{m}{n} + \Theta\left(\sqrt{\frac{m\ln n}{n}} \right)              & m \in \omega(n \ln n)
    \end{cases} \quad .
\end{align*}
\end{theorem}
\end{frame}

\begin{frame}
\frametitle{Multiple choice strategies}

\begin{exampleblock}{Idea}
Put a ball into the \alert{least full} of $d\geq2$ randomly chosen bins.
\end{exampleblock}

@@

\only<1>{
\begin{center}
\begin{tikzpicture}[scale=\scalefac]
	\bab{2 1 4 2 3 0}
	\newball
	\iA[25]{n13};
	\iA[20]{n35};
	\iA[35]{n44};
\end{tikzpicture}
\end{center}
}

@@

\begin{frame}
\frametitle{Single choice vs.~Greedy}
%\begin{columns}[onlytextwidth]
%\begin{column}{0.5\textwidth}
\begin{align*}
\loadS \leq 
	\begin{cases}
    \frac{\ln n}{\ln\ln n} + O(1)              & m = n \\
    \frac{m}{n} + \Theta\left(\sqrt{\frac{m\ln n}{n}} \right)              & m \in \omega(n \ln n)
    \end{cases}
\end{align*}
%\end{column}
%\begin{column}{0.5\textwidth}
\begin{align*}
\loadG \leq \frac{m}{n} + \frac{\ln \ln n}{\ln d} + O(1) 
\end{align*}
%\end{column}
%\end{columns}

\begin{alertblock}{The two-choice paradigm}
Even for $d=2$, Greedy achieves an \alert{exponential decrease} of the additive gap $\delta$, which does \alert{not depend on} the number of balls \alert{$m$} anymore.
\end{alertblock}
\end{frame}

\begin{frame}
\frametitle{Applications}
\begin{description}
	\item[Hashing:] Assuming a perfect hash function, $n$ elements (balls) are mapped to $m$ table entries (bins). Using two hash functions: look-up time $\Theta(\ln\ln n)$, w.h.p. \alert{Easy to parallelize} and no \alert{re-hashing}!
	\item[Online load balancing:] Assign tasks (balls) to servers/processors (bins). The two-choice paradigm helps to \alert{reduce the overhead} (of e.g., a central dispatcher), while the maximum load will be at most  $\Theta(\ln\ln n)$, w.h.p.
	\item[Emulating PRAMs on DMMs:] Distribute processors and memory cells of the PRAM on the DMM s.t.~the overhead of the emulation is minimized.
	\item[Low congestion circuit routing:] \alert{Valiant's paradigm} (a permuation routing problem can be split into two sub-problems: route to random intermediate destination and then on to the actual destination) is analogous to the single-choice technique.
\end{description}
\end{frame}

@@

\begin{center}
\begin{tikzpicture}[scale=\scalefac]
	\bab{2 1 4 2 3 0}
	\newball
	\groupSep{2};
	\groupSep{4};
	\iA[25]{n13};
	\iA[35]{n43};
	\iA[30]{n54};	
\end{tikzpicture}
\end{center}

@@

$d$-step Fibonacci numbers:
\begin{align*}
F_d(k) := \begin{cases}
                0               & k \leq 0\\
                1               & k = 1\\
                \sum_{i=1}^{d}F_d(k-i) & k > 1
            \end{cases}
\end{align*}
Generalized golden ratio $\Phi_d$:
\begin{align*}
\Phi_d := \lim_{k \rightarrow \infty} \frac{F_d(k)}{F_d(k-1)}
\end{align*}
\begin{align*}
\Phi_2 = 1.61 \quad , \quad \forall i\; \forall j \in \mathbb{N}: i < j \implies \Phi_i < \Phi_j \quad , \quad \lim_{d\rightarrow \infty} \Phi_d = 2 \quad .
\end{align*}

\begin{align*}
F_d(k) \geq \Phi_d^{k-2}
\end{align*}
\end{frame}

@@

\theoremVocking
\end{frame}

\begin{frame}
\frametitle{The Always-go-left algorithm}
\begin{columns}[onlytextwidth]
\begin{column}{0.5\textwidth}
\begin{align*}
\loadG \leq \frac{m}{n} + \frac{\ln \ln n}{\ln d} + O(1) 
\end{align*}
\end{column}
\begin{column}{0.5\textwidth}
\begin{align*}
\loadAgl \leq \frac{m}{n} + \frac{\ln \ln n}{d \ln \Phi_d} + O(1)
\end{align*}
\end{column}
\end{columns}
\bigskip
The combination of \alert{asymmetric tie breaking} and \alert{non-uniform} (and independent) sampling of the $d$ bins is crucial to achieve this upper bound.
\bigskip
\begin{theorem}[V\"ocking, 2003]
For $m=n$, using an arbitrary allocation scheme that chooses $d$ bins for each ball at random (i.e., class 3) the maximum load is, w.h.p,
\begin{align*}
\loadAgl \leq \frac{\ln \ln n}{d \ln \Phi_d} - O(1) \quad .
\end{align*}
\end{theorem}
\end{frame}

@@

% Outline of the proof section:
% - State what will be proved
% - Outline the proof
% - Witness trees 
% - Fibonacci trees and labelling
% - Lemma 3
% - Prob. of activation
\begin{frame}
\frametitle{The Always-go-left algorithm}
\framesubtitle{Proof of the upper bound for $m=n$}
\begin{theorem}[V\"ocking, 2003]
\label{theorem:aglm}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Always-go-left algorithm, the maximum load at any time step $t$ is, w.h.p., 
\begin{align*}
\loadAgl \leq \frac{\ln\ln n}{d  \ln \Phi_d} + O(1) \quad .
\end{align*}
\end{theorem}

\bigskip
In case $\load$ exceeds a given threshold of $L+3$ balls ("a bad event occurs"), we construct an activated \alert{witness tree of order $L$}. The probability for its existance upper bounds the probabilty that the associated bad event occurs.  
\end{frame}

@@

\bigskip

\assumptionOne

I.e., we assume that a ball is referenced by at most one node.
\end{frame}

@@

\begin{center}
\begin{tikzpicture}
\matrix{
 \Tree 
	[.$T_d(6)$
		[.$T_d(5)$
			[.$T_d(4)$ 
				[.$T_d(3)$ $T_d(2)$ $T_d(1)$ ]
				$T_d(2)$ $T_d(1)$
			]
			[.$T_d(3)$ $T_d(2)$ $T_d(1)$ ]
			$T_d(2)$ %$T_d(1)$
		]
		[.$T_d(4)$ 
			[.$T_d(3)$ $T_d(2)$ $T_d(1)$ ]
			$T_d(2)$ $T_d(1)$
		]
		[.$T_d(3)$ $T_d(2)$ $T_d(1)$ ]
		%$T_d(2)$ 
	]
\\};
\end{tikzpicture}
\end{center}
\end{frame}

@@

\begin{frame}
\frametitle{Edge and leaf events}
\begin{columns}[onlytextwidth]
\begin{column}{0.7\textwidth}
\defEdgeEvent

\defLeafEvent
\end{column}
\begin{column}{0.3\textwidth}
\begin{tikzpicture}
\todo{illustrate}
\end{tikzpicture}
\end{column}
\end{columns}

@@

\begin{frame}[shrink]
\frametitle{Asymmetric witness trees (AWT)} 
We assign a label $(h,i)$, with $0\leq h \leq L$, $1\leq i \leq d$, to each node:
\begin{itemize}
\item Node $T_d(1)$/$T_d(2)$ is labelled $(0,1)$/$(0,2)$.
\item A node labelled $(0,i)$, $i>2$, has $i-1$ children labelled $(0,i-1),\dots,(0,1)$.
\item A node with label $(h,i)$, $h>0$, has $d$ children labelled
$(h,i-1),\dots,(h,1)$,$(h-1,d),\dots,(h-1, i)$.
\item Thus, the node $T_d(dh+i)$ is labelled $(h,i)$.
\end{itemize}
\begin{center}
\begin{tikzpicture}
\matrix{
 \Tree 
	[.$(1,2)$ 
		[.$(1,1)$
			[.$(0,3)$ $(0,2)$ $(0,1)$ ]
			$(0,2)$ $(0,1)$
		]
		[.$(0,3)$ $(0,2)$ $(0,1)$ ]
		$(0,2)$
	]
\\};
\end{tikzpicture}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Proof of activation}
We assign balls to the nodes of the AWT s.t.~th

\begin{exampleblock}{Invariant}
A ball $b_v$ referenced by a node $v$ with label $(h,i)$ was placed in a bin $x$ belonging to group $i$ that, at the time of insertion of the ball $b_v$, contained at least $h+3$ other balls. 
\end{exampleblock}

@@

\begin{frame}
\frametitle{Probability of activation}
\begin{lemma}
For any $\alpha > 0$ and large enough $L$, the activation of an AWT of order L occurs with probability at most $n^{-\alpha}$.
\end{lemma}

\bigskip
\begin{itemize}
\item $k$\dots number of nodes
\item $l$\dots number of leaves
\item $\omega$\dots number of ways to assign balls to the nodes of an AWT
\item $p_\mathrm{e}$\dots probability that all edge events occur
\item $p_\mathrm{l}$\dots probability that all edge events occur
\item $P_\mathrm{a}$\dots probability of activation of an AWT of order $L$
\end{itemize}

@@

\begin{frame}
\frametitle{Leaf events}
\only<1>{
\defLeafEvent
At any time step $t$, at most $n/3$ bins contain 3 or more balls. \alert{But how are they distributed among the $d$ groups?}

\medskip
$b_i$\dots number of bins in group $i$ containing 3 or more balls.
$\beta_i = b_i d/n$\dots fraction of bins $b_i$ in group $i$. 
}
\only<1-2>{
\begin{align*}
\frac{n}{d}\sum_{i=1}^d \beta_i \leq \frac{n}{3} \implies
\sum_{i=1}^d \beta_i \leq \frac{d}{3}
\end{align*}
}

@@

\begin{frame}
\frametitle{Probability of activation}
Putting it together, we get
\begin{align*}
P_\mathrm{a} = \omega p_e p_l \leq  n^k \left(\frac{d}{n}\right)^{k-1} 3^{-dl} \quad .
\end{align*}
Using $ k\leq 2l$, $2d^2\leq 3^d$ and $l \geq \Phi_d^{dL-1}$, we can simplify to
\begin{align*}
P_\mathrm{a} \leq n 2 ^{-\Phi_d^{dL-1}} \quad .
\end{align*}
For arbitrary $\alpha > 0$, let
\begin{align*}
L &\geq \left\lceil{\frac{\ln\log_2 n + \ln\left(1+\alpha\right)}{d \ln \Phi_d}}\right\rceil+\frac{1}{d} \implies \frac{\ln\ln n}{d \ln \Phi_d} + O\left(1\right) ,
\end{align*}
and we get
\begin{align*}
P_\mathrm{a} \leq n^{-\alpha} \quad  \qed
\end{align*}
\end{frame}

\begin{frame}
\frametitle{Recap.}
\theoremVocking
\assumptionOne
\bigskip 
Next steps:
\begin{itemize}
\item Proof \alert{without} the simplifying assumption (V\"ocking, 2003).
\item Proof for the \alert{heavily loaded} case (Berenbrink et al.,~2006).
\end{itemize}

\begin{frame}
\frametitle{Conclusion}
\end{frame}

\begin{frame}[allowframebreaks]
\frametitle<presentation>{References}    
\bibliographystyle{acm}
\nocite{VOC03}
\nocite{ABKU99}
\nocite{RS98}
\nocite{BCSV06}
\nocite{capocelli1990generalized}
\setbeamertemplate{bibliography item}[text]
\bibliography{../sources}
\end{frame}

>>>>

\newcommand\xqed[1]{%
   \rlap{\hbox to#1{\hfil\llap{\ensuremath{\qed}}}}
}

@@

%---------------------------------------------------------------------
%parameters for balls into bins graphs
%---------------------------------------------------------------------
%scale factor
\newcommand\scalefac{0.55}
%diameter of balls
\newcommand\ballsize{5mm}
%number of bins
\newcommand\nrbins{6}
%space between ball and bin
\newcommand\padding{0.1*\ballsize}
%height of a bin
\newcommand\binheight{5*\balldiameter}
%width of a bin
\newcommand\binwidth{\balldiameter}
%gap from bin to bin
\newcommand\bingap{1.6*\balldiameter}
%color of balls
\tikzstyle{ballstyle} = [ball color=black!30!red]
\tikzstyle{specialBallstyle} = [ball color=black!30!red,  opacity=0.2]

@@

%paint a ball in bin #1, position #2 with style specialBallstyle
\newcommand\specialBall[2]{
	\shade[specialBallstyle] (#1*\bingap,#2*\balldiameter-\ballsize) circle (\ballsize) {};
}

%paint a ball in bin #1, position #2 with style = ballstyle
\newcommand\ball[2]{
	\shade[ballstyle] (#1*\bingap,#2*\balldiameter-\ballsize) circle (\ballsize) {};
}

@@

% nicer tikz qtrees
\tikzset{edge from parent/.style=
     {draw, edge from parent path={(\tikzparentnode) -- (\tikzchildnode)}}}
\tikzset{every tree node/.style={font=\tiny, thick, align=center}}

%---------------------------------------------------------------------
%Front matter
%---------------------------------------------------------------------
\title{The Power of Two Random Choices}
\subtitle{And the Surprising Influence of Asymmetry}
\author[M. Kalany]{Martin Kalany}
\institute[TU Wien]
{
  Graduate student in Computer Science\\
  Vienna University of Technology\\
}
\date{\today}

%---------------------------------------------------------------------
%document
%---------------------------------------------------------------------
% Structure:
% 30 min
% 5 min easy stuff (so that everyone understands)
% 5 min more advanced stuff 
% 15 min difficult stuff (the proof, of the asymmetric allocation scheme using the infamous assumption 1)
% 5 min further results and wrap up
\begin{document}
\begin{frame}
  \titlepage
\end{frame}

\begin{frame}
\frametitle{The problem}
\framesubtitle{...and the goal}
\begin{exampleblock}{Balls-into-bins games}
Suppose we have $n$ initially empty bins and need to distribute $m \geq n$ balls among them, \alert{as evenly as possible}. 
\end{exampleblock}
\pause
\begin{exampleblock}{Definition}
We are interested in the \alert{maximum load $\load$}, i.e, how many balls are in the fullest bin?
\end{exampleblock}

@@

\begin{frame}
\frametitle{Single choice strategy}
\framesubtitle{Upper bound for the maximum load}
%&Since we deal with a \alert{random process}, we cannot give upper bounds for the maximum load $\load$ with absolute certainty!

@@

\pause
\medskip
We distinguish the \alert{lightly loaded} case ($m = n$) and the \alert{heavily loaded} case ($m \in \omega(n\ln n)$).

\medskip
\begin{theorem}[Raab and Steger, 1998]
When using the single choice strategy to distribute $m$ balls to $n$ bins, the maximum load will be, w.h.p.,
\begin{align*}
\loadS \leq 
	\begin{cases}
    \Theta\left(\frac{\ln n}{\ln\ln n}\right)           & m = n \\
    \frac{m}{n} + \Theta\left(\sqrt{\frac{m\ln n}{n}} \right)              & m \in \omega(n \ln n)
    \end{cases} \quad .
\end{align*}
\end{theorem}

\begin{frame}
\frametitle{Multiple choice strategies}
\begin{exampleblock}{Idea}
Put a ball into the \alert{least full} of $d\geq2$ randomly chosen bins.
\end{exampleblock}

@@

\only<1>{
\begin{center}
\begin{tikzpicture}[scale=\scalefac]
	\bab{2 1 4 2 3 0}
	\newball
	\iA[35]{n22};
	\iA[35]{n44};
\end{tikzpicture}
\end{center}
}

@@

\begin{frame}
\frametitle{Single choice vs.~Greedy}
\begin{align*}
\loadS \leq 
	\begin{cases}
    \frac{\ln n}{\ln\ln n} + O(1)              & m = n \\
    \frac{m}{n} + \Theta\left(\sqrt{\frac{m\ln n}{n}} \right)              & m \in \omega(n \ln n)
    \end{cases}
\end{align*}
\begin{align*}
\loadG \leq \frac{m}{n} + \frac{\ln \ln n}{\ln d} + O(1) 
\end{align*}

\begin{alertblock}{The two-choice paradigm}
Even for $d=2$, Greedy achieves an \alert{exponential decrease} of the additive gap $\delta$, which does \alert{not depend on} the number of balls \alert{$m$}.
\end{alertblock}
\end{frame}

\begin{frame}
\frametitle{Applications}
\begin{description}
	\item[Hashing:] Assuming a perfect hash function, $n$ elements (balls) are mapped to $m$ table entries (bins). Using two hash functions: look-up time $\Theta(\ln\ln n)$, w.h.p. \alert{Easy to parallelize} and  \alert{no re-hashing}!
	\item[Online load balancing:] Assign tasks (balls) to servers/processors (bins). The two-choice paradigm helps to \alert{reduce communication overhead}, %(of e.g., a central dispatcher)
	while the maximum load will be $\Theta(\ln\ln n)$, w.h.p.
	\item[Emulating PRAMs on DMMs:] Distribute processors and memory cells of the PRAM on the DMM s.t.~the overhead of the emulation is minimized.
	\item[Low congestion circuit routing:] \alert{Valiant's paradigm} %(a permuation routing problem can be split into two sub-problems: route to random intermediate destination and then on to the actual destination) 
	is analogous to the single-choice technique.
\end{description}
\end{frame}

@@

\begin{center}
\begin{tikzpicture}[scale=\scalefac]
	\bab{2 1 4 2 3 0}
	\newball
	\groupSep{3};
	\iA[30]{n13};
	\iA[30]{n54};	
\end{tikzpicture}
\end{center}

@@

$d$-step Fibonacci numbers:
\begin{align*}
F_d(k) := \begin{cases}
                0               & k \leq 0\\
                1               & k = 1\\
                \sum_{i=1}^{d}F_d(k-i) & k > 1
            \end{cases}
\end{align*}
Generalized golden ratio $\Phi_d$:
\begin{align*}
\Phi_d := \lim_{k \rightarrow \infty} \frac{F_d(k)}{F_d(k-1)}
\end{align*}
\begin{align*}
\Phi_2 = 1.61 \quad , \quad \forall i\; \forall j \in \mathbb{N}: i < j \implies \Phi_i < \Phi_j \quad , \quad \lim_{d\rightarrow \infty} \Phi_d = 2 \quad .
\end{align*}
\begin{align*}
F_d(k) \geq \Phi_d^{k-2}
\end{align*}
\end{frame}

@@

\begin{frame}[shrink]
\frametitle{The Always-go-left algorithm}
\aglAlgorithm
\theoremVocking
\end{frame}

\begin{frame}
\frametitle{The Always-go-left algorithm}
\begin{columns}[onlytextwidth]
\begin{column}{0.5\textwidth}
\begin{align*}
\loadG \leq \frac{m}{n} + \frac{\ln \ln n}{\ln d} + O(1) 
\end{align*}
\end{column}
\begin{column}{0.5\textwidth}
\begin{align*}
\loadAgl \leq \frac{m}{n} + \frac{\ln \ln n}{d \ln \Phi_d} + O(1)
\end{align*}
\end{column}
\end{columns}
\bigskip
The combination of \alert{asymmetric tie breaking} and \alert{non-uniform} (and independent) sampling of the $d$ bins is crucial to achieve this upper bound.
\bigskip
\begin{theorem}[V\"ocking, 2003]
For $m=n$, using an arbitrary allocation scheme that chooses $d$ bins for each ball at random (i.e., class 3) the maximum load is, w.h.p,
\begin{align*}
\loadAgl = \frac{\ln \ln n}{d \ln \Phi_d} - O(1) \quad .
\end{align*}
\end{theorem}
\end{frame}

@@

\begin{frame}
\frametitle{The Always-go-left algorithm}
\framesubtitle{Proof of the upper bound for $m=n$}
\begin{theorem}[V\"ocking, 2003]
\label{theorem:aglm}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Always-go-left algorithm, the maximum load at any time step $t$ is, w.h.p., 
\begin{align*}
\loadAgl \leq \frac{\ln\ln n}{d  \ln \Phi_d} + O(1) \quad .
\end{align*}
\end{theorem}

\pause
\bigskip
In case $\load$ exceeds a given threshold of $L+3$ balls ("a bad event occurs"), we construct an activated \alert{witness tree of order $L$}. The probability for its existence upper bounds the probability that the associated bad event occurs.  
\end{frame}

@@

\bigskip
\assumptionOne
\end{frame}

@@

\begin{center}
\begin{tikzpicture}[scale=1]
 \Tree 
 	[.$T_d(7)$ \edge[draw, dotted];
		[.$T_d(6)$
			[.$T_d(5)$
				[.$T_d(4)$ 
					[.$T_d(3)$ $T_d(2)$ $T_d(1)$ ]
					$T_d(2)$ %$T_d(1)$
				]
				[.$T_d(3)$ $T_d(2)$ $T_d(1)$ ]
				%$T_d(2)$
			]
			[.$T_d(4)$ 
				[.$T_d(3)$ $T_d(2)$ $T_d(1)$ ]
				$T_d(2)$ %$T_d(1)$
			]
			%[.$T_d(3)$ $T_d(2)$ $T_d(1)$ ]
		]
		\edge[draw=none]; {}
		%\edge[draw=none]; {}		
	]
\end{tikzpicture}
\end{center}
\end{frame}

@@

\begin{frame}
\frametitle{Edge events}
\defEdgeEvent
\begin{center}
\begin{tikzpicture}[every tree node/.style={thick}]
 \Tree 
 	[.\node[draw,circle]{u}; 
		\edge[draw, densely dotted]; {}  
		%\edge[draw, densely dotted]; {}
		\edge[draw] node[draw=none, thick, anchor=west] {$i$};
		[.\node[draw,circle]{v};
		]
	]
	\node (bu) at (0.75,0.5) {};
	\node (bv) at (1.1,-1.25) {};

	\path node[topflat, xshift=3.5cm, above, yshift=-3cm]  {};
	\node[circle] (bin1) at (3.5,-0.75) {};

	\path node[topflat, xshift=4.5cm, above, yshift=-3cm]  {};
	\node[circle] (bin2) at (4.5,-0.75) {};

	\path node[topflat, xshift=5.5cm, above, yshift=-3cm]  {};
	\node[circle] (bin3) at (5.5,-0.75) {};

	\path node[topflat, xshift=6.5cm, above, yshift=-3cm]  {};
	\node[circle] (bin4) at (6.5,-0.75) {};

	\path node[topflat, xshift=7.5cm, above, yshift=-3cm]  {};
	\node[circle] (bin5) at (7.5,-0.75) {};

	\path node[topflat, xshift=8.5cm, above, yshift=-3cm]  {};
	\node[circle] (bin6) at (8.5,-0.75) {};

	\path (bu) edge[insert, bend left=20] node[draw=none, font=\tiny, sloped, anchor=south]{$i$-th location} (bin3);
	\path (bu) edge[insert, dotted, bend left=30] (bin6);

	\path (bv) edge[insert, bend right=30] node[draw=none, font=\tiny, sloped, anchor=south]{any location} (bin3);
	\path (bv) edge[insert, dotted, bend right=40] (bin4);

	\shade[ballstyle] (0.75,0.5) circle (\ballsize*\scalefac) {};
	\shade[ballstyle] (1.1,-1.25) circle (\ballsize*\scalefac) {};
\end{tikzpicture}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Leaf events}
\defLeafEvent

\bigskip
\begin{center}
\begin{tikzpicture}[every tree node/.style={thick}]
\begin{scope}[xshift=-0.9cm, yshift=4cm]
\Tree
 	[.\node[draw,circle]{u}; 
		\edge[draw, densely dotted]; {}  
		\edge[draw] {};
		[.\node[draw,circle]{v};
		]
	]
\end{scope}
\begin{tikzpicture}[scale=\scalefac]
\begin{scope}[yshift=1cm]
	\bab{3 0 0 0 3 0}
	\newball
	\iA[30]{n14};
	\iA[25]{n54};	
\end{scope}
\end{tikzpicture}

\end{tikzpicture}
\end{center}

@@

\begin{frame}[shrink]
\frametitle{Asymmetric witness trees (AWT)} 
We assign a label $(h,i)$, with $0\leq h \leq L$, $1\leq i \leq d$, to each node:
\begin{itemize}
\item Node $T_d(1)$/$T_d(2)$ is labeled $(0,1)$/$(0,2)$.
\item For $i>2$, a node $(0,i)$ has $i-1$ children $(0,i-1),\dots,(0,1)$.
\item For $h>0$, a node $(h,i)$, has $d$ children
$(h,i-1),\dots,(h,1)$, $(h-1,d),\dots,(h-1, i)$.
\item Thus, the node $T_d(dh+i)$ is labeled $(h,i)$.
\end{itemize}
\begin{center}
\begin{tikzpicture}
 \Tree 
	[.$T_d(6)$\\$(1,3)$ \edge[draw, dotted];
		[.$T_d(5)$\\$(1,2)$
			[.$T_d(4)$\\$(1,1)$
				[.$T_d(3)$\\$(0,3)$ $T_d(2)$\\$(0,2)$ $T_d(1)$\\$(0,1)$ ]
				$T_d(2)$\\$(0,2)$ %$T_d(1)$\\$(0,1)$
			]
			[.$T_d(3)$\\$(0,3)$ $T_d(2)$\\$(0,2)$ $T_d(1)$\\$(0,1)$ ]
			%$T_d(2)$\\$(0,2)$
		]
	\edge[draw=none]; {}
	]
\end{tikzpicture}
\end{center}
\end{frame}

\newcommand\myInvariant{
\begin{exampleblock}{Invariant}
A ball $b_v$ referenced by a node $v$ with label $(h,i)$ was placed in a bin $x$ belonging to group $i$ that, at the time of insertion of the ball $b_v$, contained at least $h+3$ other balls. 
\end{exampleblock}
}

\begin{frame}
\frametitle{Proof of activation}
\begin{block}{Lemma}
The occurrence of a bad event (i.e., a bin contains more than $L+3$) implies the activation of an AWT of order $L$.
\end{block}
\bigskip
\visible<2>{
We assign balls to the nodes of the AWT s.t.:
\myInvariant
}
\end{frame}

\begin{frame}
\frametitle{Proof of activation}

\myInvariant

Let a bin contain $L+4$ balls. We assign the topmost ball to the root of the AWT:
\begin{columns}[onlytextwidth]
\begin{column}{0.5\textwidth}
\begin{center}
\begin{tikzpicture}[scale=1, edge from parent/.style=
     {draw, edge from parent path={(\tikzparentnode) -- (\tikzchildnode)}}]
\Tree 
	[.$(L,1)$ \edge[draw=black!30!red, thick];
		[.$(L-1,3)$
			[.$(L-1,2)$
				[.$(L-1,1)$ ]
				\edge[draw, densely dotted]; {}
			]
			\edge[draw, densely dotted]; {}
		]
		\edge[draw=black!30!red, thick];
		[.$(L-1,2)$ \edge[draw, densely dotted]; {} \edge[draw, densely dotted]; {} ]
	]
\end{tikzpicture}
\end{center}
\end{column}
\begin{column}{0.5\textwidth}
\begin{tikzpicture}[scale=\scalefac]
	\bab{2 4 4 2 3 4}
	\groupSep{3};
	\setNode{1}{7};
	\ball{1}{7};
	\specialBall{2}{5};
	\path (n17) edge[insert, bend left=25] (n25);
	\path (n17) edge[insert, bend left=25] (n65);
\end{tikzpicture}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Proof of activation}
Given a node $v$ with label $(h,i)$ that references a ball $b_v$ (which was put into some bin of group $i$ that contained at least $h+3$ other balls at the time of insertion of ball $b_v$), \alert{we conclude} that:
\begin{enumerate}
\only<1>{
\item For $1\leq j < i$, the $j$-th location of the ball $b_v$ contains at least $h+4$ balls, at the insertion time of the ball $b_v$. If it contained less balls than the $i$-th location, the Always-go-left scheme would have put the ball $b_v$ into the bin pointed to by location $j$. We assign the topmost ball in location $j$ to the child $u$ of $v$ with label $(h,j)$. 
}
\only<2>{
\item For $i < j \leq d$, the $j$-th location contains at least $h+3$ balls. If it contained less, the Always-go-left scheme would have put the ball $b_v$ into the bin location $j$ points to. We assign the topmost ball in location $j$ to the child node $u$ of $v$ labeled with $(h-1, j)$. \qed

}
\end{enumerate}

@@

\begin{frame}
\frametitle{Probability of activation}
\begin{lemma}
For any $\alpha > 0$ and large enough $L$, the activation of an AWT of order L occurs with probability at most $n^{-\alpha}$.
\end{lemma}
\medskip
\begin{itemize}
\item $k$\dots number of nodes
\item $l$\dots number of leaves
\item $\omega$\dots number of ways to assign balls to the nodes of an AWT
\item $p_\mathrm{e}$\dots probability that all edge events occur
\item $p_\mathrm{l}$\dots probability that all edge events occur
\item $P_\mathrm{a}$\dots probability of activation of an AWT of order $L$
\end{itemize}

@@

\begin{frame}
\frametitle{Leaf events}
\only<1>{
\defLeafEvent
At any time step $t$, at most $n/3$ bins contain 3 or more balls. But \alert{how are they distributed} among the $d$ groups?

\medskip
$b_i$\dots number of bins in group $i$ containing 3 or more balls
$\beta_i = b_i d/n$\dots fraction of bins $b_i$ in group $i$
}
\only<1-2>{
\begin{align*}
\frac{n}{d}\sum_{i=1}^d \beta_i \leq \frac{n}{3} \implies
\sum_{i=1}^d \beta_i \leq \frac{d}{3}
\end{align*}
}

@@

\begin{frame}
\frametitle{Probability of activation}
Putting it together, we get
\begin{align*}
P_\mathrm{a} = \omega p_e p_l \leq  n^k \left(\frac{d}{n}\right)^{k-1} 3^{-dl} \quad .
\end{align*}
Using $ k\leq 2l$, $2d^2\leq 3^d$ and $l \geq \Phi_d^{dL-1}$, we can simplify to
\begin{align*}
P_\mathrm{a} \leq n 2 ^{-\Phi_d^{dL-1}} \quad .
\end{align*}
For arbitrary $\alpha > 0$, let
\begin{align*}
L &\geq \left\lceil{\frac{\ln\log_2 n + \ln\left(1+\alpha\right)}{d \ln \Phi_d}}\right\rceil+\frac{1}{d} \implies \frac{\ln\ln n}{d \ln \Phi_d} + O\left(1\right) ,
\end{align*}
and we get
\begin{align*}
P_\mathrm{a} \leq n^{-\alpha} \xqed{5cm}
\end{align*}
\end{frame}

\begin{frame}
\frametitle{The big picture}
\theoremVocking
\assumptionOne
\medskip 
Next steps:
\begin{itemize}
\item Prove \alert{without} the simplifying assumption (V\"ocking, 2003).
\item Prove for the \alert{heavily loaded} case (Berenbrink et al.,~2006).
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Conclusion}
\begin{enumerate}
\item The balls-into-bins model \alert{adopts naturally} to various problems (hashing, load balancing,\dots).
\item Applying the two-choice paradigm reduces the upper bound for the maximum load by an \alert{exponential factor} (compared to a single choice scheme).
\item \alert{Many variations} possible: Non-uniform balls/bins; weighted balls; parallel allocation of several balls;\dots
\item \alert{Witness trees} are a suitable proof technique for this kind of problems.
\end{enumerate}
\end{frame}

\begin{frame}[shrink]
\frametitle<presentation>{References}    
\bibliographystyle{acm}
\nocite{VOC03}
\nocite{ABKU99}
\nocite{RS98}
\nocite{BCSV06}
\nocite{capocelli1990generalized}
\setbeamertemplate{bibliography item}[text]
\bibliography{../sources}
\end{frame}
